\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,booktabs}
\usepackage[margin=1in]{geometry}

\title{MVP Simulator: MDP Formulation and Initial Results}
\author{William Dennis}
\date{June 2026}

\begin{document}
\maketitle

\section{MDP Formulation}

The MVP simulator implements a finite-horizon MDP $(\mathcal{S}, \mathcal{A}, P, r, \gamma, T)$:

\subsection{State Space}
\[
\mathcal{S} = \{\texttt{sedentary},\ \texttt{active}\} \times \mathcal{T}
\]
where $\mathcal{T} = \{\texttt{morning}, \texttt{midday}, \texttt{afternoon}, \texttt{evening}, \texttt{night}\}$ are the five daily decision points.

\subsection{Action Space}
\[
\mathcal{A} = \{\texttt{send},\ \texttt{don\_t\_send}\}
\]
A binary intervention: send a motivational message or do not.

\subsection{Transition Function}
$P(s' \mid s, a, t)$ is defined by a configurable matrix:

\begin{equation}
P(\texttt{active} \mid \texttt{sedentary}, \texttt{send}) = 0.3, \quad
P(\texttt{active} \mid \texttt{sedentary}, \texttt{don\_t\_send}) = 0.1
\end{equation}

\begin{equation}
P(\texttt{active} \mid \texttt{active}, \texttt{send}) = 0.8, \quad
P(\texttt{active} \mid \texttt{active}, \texttt{don\_t\_send}) = 0.6
\end{equation}

Time-of-day mask: at \texttt{night}, all transitions are suppressed (availability indicator per HeartSteps V2).

\subsection{Reward}
\[
r(s, a) =
\begin{cases}
1.0 & \text{if } s'.\texttt{activity} = \texttt{active} \\
0.0 & \text{otherwise}
\end{cases}
\]
Single-timescale placeholder. Multi-timescale reward (immediate + delayed body measure) is deferred to Phase 2.

\subsection{Episode Structure}
$T = 90 \text{ days} \times 5 \text{ steps/day} = 450 \text{ timesteps}$.

\section{Agents}

All agents are contextual bandits: state is accepted but not used in action selection. For the stationary MVP transition matrix, $E[r \mid a]$ is the relevant quantity.

\subsection{Thompson Sampling}
Beta-Bernoulli TS with prior $\text{Beta}(1, 1)$ per action. Posterior update:
\[
\alpha_a \leftarrow \alpha_a + \mathbb{1}[s'.\texttt{activity} = \texttt{active}], \quad
\beta_a \leftarrow \beta_a + \mathbb{1}[s'.\texttt{activity} = \texttt{sedentary}]
\]
Action selected by sampling $\theta_a \sim \text{Beta}(\alpha_a, \beta_a)$ and choosing $a^* = \arg\max_a \theta_a$.

\subsection{Epsilon-Greedy}
$\epsilon = 0.1$. With probability $\epsilon$, select uniformly at random; otherwise select $\arg\max_a Q(a)$. Q-values updated via incremental average:
\[
Q(a) \leftarrow Q(a) + \frac{r_t - Q(a)}{n_a}
\]

\subsection{Random Baseline}
Uniform random action selection. No learning. Serves as lower bound.

\section{Initial Results}

All agents run on the same config (\texttt{config/mvp.yaml}), seed 42, 450 timesteps.

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Agent & Total Reward & Mean Reward/Step \\
\midrule
Thompson Sampling & 252.0 & 0.5600 \\
Epsilon-Greedy ($\epsilon=0.1$) & 253.0 & 0.5622 \\
Random & 182.0 & 0.4044 \\
\bottomrule
\end{tabular}
\caption{Results on \texttt{config/mvp.yaml} (seed 42, 450 timesteps). TS and $\epsilon$-greedy both learn the optimal policy; random serves as lower bound.}
\end{table}

\noindent\textbf{Note:} These are placeholder results from the CLI smoke test. Full multi-agent comparison with seeded reproducibility will be added in a follow-up.

\section{Known Limitations}

\begin{itemize}
\item Agents are state-blind bandits, not state-aware RL. State-aware agents are Phase 2.
\item Single-timescale reward only. Multi-timescale (immediate + 3-week delayed) is Phase 2.
\item No user response model. The \texttt{RuleBasedResponse} module exists as a stub. Burden accumulation and 4 archetypes are Phase 2.
\item Transition matrix is fixed and known. Learned transitions from real data are Phase 2.
\end{itemize}

\end{document}
