\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,booktabs}
\usepackage[margin=1in]{geometry}

\title{MVP Simulator: MDP Formulation and Initial Results}
\author{William Dennis}
\date{June 2026}

\begin{document}
\maketitle

\section{MDP Formulation}

The MVP simulator implements a finite-horizon MDP $(\mathcal{S}, \mathcal{A}, P, r, \gamma, T)$:

\subsection{State Space}
\[
\mathcal{S} = \{\texttt{sedentary},\ \texttt{active}\}
\]
Two activity states with no additional time-of-day dimension. Time-dependent reward scaling is handled via \texttt{reward\_multiplier\_by\_step} (see \S\ref{sec:reward} below).

\subsection{Action Space}
\[
\mathcal{A} = \{\texttt{nudge},\ \texttt{idle}\}
\]
A binary intervention: send a motivational message (\texttt{nudge}) or do not (\texttt{idle}).

\subsection{Transition Function}

$P(s' \mid s, a)$ is a configurable $2 \times 2 \times 2$ tensor. Rows index current state $s \in \{\text{sedentary}, \text{active}\}$ and columns index next state $s'$ in the same order:

\[
P_{\text{nudge}} =
\begin{bmatrix}
0.7 & 0.3 \\[2pt]
0.5 & 0.5
\end{bmatrix}
\qquad
P_{\text{idle}} =
\begin{bmatrix}
0.9 & 0.1 \\[2pt]
0.4 & 0.6
\end{bmatrix}
\]

Sending a reminder increases $P(\text{active} \mid \text{sedentary})$ from $0.1$ to $0.3$ but decreases $P(\text{active} \mid \text{active})$ from $0.6$ to $0.5$, consistent with Reactance theory (reminders can backfire for already-active users).

Optional \texttt{reward\_multiplier\_by\_step}: a per-step scaling factor that can zero out rewards at certain daily steps (soft mask). When absent, all multipliers default to 1.0. The per-step reward is computed as $\text{base reward} \times \text{multiplier}[t]$ where $t$ is the step-of-day index.

\subsection{Reward}\label{sec:reward}

The base reward is a vector indexed by next state:

\[
r = [0.0,\; 1.0]^\top \quad (\text{sedentary},\; \text{active})
\]

The per-step reward at time $t$ scales by the multiplier:

\[
r(s, a, t) = m[t \bmod S] \cdot r(s')
\]

where $S = \texttt{steps\_per\_day}$, $m \in \mathbb{R}^S$ is \texttt{reward\_multiplier\_by\_step}
(default $m_i = 1.0$), and $s'$ is the next state after taking action $a$ in state $s$.
Currently $m_i = 1.0$ for all $i$, so $r(s, a, t) = r(s')$.

Single-timescale placeholder. Multi-timescale reward (immediate + delayed body measure) is deferred to Phase 2.

\subsection{Episode Structure}
$T = \texttt{episode\_days} \times \texttt{steps\_per\_day}$ timesteps. Default: $T = 90 \text{ days} \times 5 \text{ steps/day} = 450$.

\section{Agents}

All agents are bandits: they treat each decision step as an independent lever-pull and do not condition on the state.
Because the transition matrix is stationary, each action $a$ induces a Markov chain whose
stationary distribution $\pi_a$ follows from $P_a$:

\[
\pi_a(\text{active}) = \frac{P(\text{active} \mid \text{sedentary}, a)}
{P(\text{active} \mid \text{sedentary}, a) + P(\text{sedentary} \mid \text{active}, a)},
\qquad
\pi_a(\text{sedentary}) = 1 - \pi_a(\text{active})
\]

The expected reward for always choosing action $a$ is:

\[
E[r \mid a] = \sum_{s'} \pi_a(s') \, r(s') = \pi_a(\text{active})
\]

From the MVP matrices:

\[
E[r \mid \text{nudge}] = \frac{0.3}{0.3 + 0.5} \approx 0.375, \qquad
E[r \mid \text{idle}] = \frac{0.1}{0.1 + 0.4} = 0.20
\]

Nudge dominates idle in expectation, but a state-aware agent could improve further by
conditioning on $s$ (nudge when sedentary, idle when active), yielding
$\pi(\text{active}) \approx 0.429$. This limitation motivates the state-aware agents
planned for Phase~2.

\subsection{Thompson Sampling}
Beta-Bernoulli TS with prior $\text{Beta}(1, 1)$ per action. Posterior update:
\[
\alpha_a \leftarrow \alpha_a + \mathbb{1}[s'.\texttt{activity} = \texttt{active}], \quad
\beta_a \leftarrow \beta_a + \mathbb{1}[s'.\texttt{activity} = \texttt{sedentary}]
\]
Action selected by sampling $\theta_a \sim \text{Beta}(\alpha_a, \beta_a)$ and choosing $a^* = \arg\max_a \theta_a$.

\subsection{Epsilon-Greedy}
$\epsilon = 0.1$. With probability $\epsilon$, select uniformly at random; otherwise select $\arg\max_a Q(a)$. Q-values updated via incremental average:
\[
Q(a) \leftarrow Q(a) + \frac{r_t - Q(a)}{n_a}
\]

\subsection{UCB1}
Upper Confidence Bound with exploration parameter $c = 2.0$. Selects:
\[
a^* = \arg\max_a \left[ Q(a) + c \sqrt{\frac{\ln N}{N_a}} \right]
\]
where $N$ is total steps and $N_a$ is the number of times action $a$ was chosen. The confidence bonus shrinks as $N_a$ grows, naturally shifting from exploration to exploitation. Each action is pulled once before UCB is applied.

\subsection{Random Baseline}
Uniform random action selection. No learning. Serves as lower bound.

\section{Initial Results}

All agents run over 10 random seeds (1--10), 450 timesteps each. Results are reported as mean $\pm$ standard deviation across seeds.

\subsection{Standard Config (\texttt{config/rule\_based.yaml})}
No \texttt{reward\_multiplier\_by\_step} -- uniform 1.0 multiplier on all steps.

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Agent & Total Reward & Mean Reward/Step \\
\midrule
Thompson Sampling ($\alpha_0=\beta_0=1$) & 160.3 $\pm$ 10.5 & 0.356 $\pm$ 0.023 \\
Epsilon-Greedy ($\epsilon=0.1$) & 160.8 $\pm$ 15.3 & 0.357 $\pm$ 0.034 \\
UCB ($c=2.0$) & 148.2 $\pm$ 15.0 & 0.329 $\pm$ 0.033 \\
Random & 135.6 $\pm$ 10.7 & 0.301 $\pm$ 0.024 \\
\bottomrule
\end{tabular}
\caption{Results on \texttt{config/rule\_based.yaml} (10 seeds, 450 timesteps). TS and $\epsilon$-greedy perform similarly; TS has lower variance. UCB outperforms random but shows higher variance.}
\end{table}

\subsection{Masked Config (\texttt{config/rule\_based\_with\_mask.yaml})}
\texttt{reward\_multiplier\_by\_step: [1, 1, 1, 1, 0]} -- step 4 (end of day) yields zero reward regardless of outcome. All agents drop proportionally to the masked fraction ($1/5$ of steps zeroed), confirming the multiplicative reward model.

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Agent & Total Reward & Mean Reward/Step \\
\midrule
Thompson Sampling ($\alpha_0=\beta_0=1$) & 126.6 $\pm$ 6.8 & 0.281 $\pm$ 0.015 \\
Epsilon-Greedy ($\epsilon=0.1$) & 127.3 $\pm$ 10.2 & 0.283 $\pm$ 0.023 \\
UCB ($c=2.0$) & 116.9 $\pm$ 8.6 & 0.260 $\pm$ 0.019 \\
Random & 107.8 $\pm$ 6.7 & 0.240 $\pm$ 0.015 \\
\bottomrule
\end{tabular}
\caption{Results on \texttt{config/rule\_based\_with\_mask.yaml} (10 seeds, 450 timesteps). Reward drops $\approx 20\%$ across all agents, matching the $1/5$ masked fraction. Relative ordering is preserved.}
\end{table}

\section{Known Limitations}

\begin{itemize}
\item Agents are state-blind bandits, not state-aware RL. State-aware agents are Phase 2.
\item Single-timescale reward only. Multi-timescale (immediate + 3-week delayed) is Phase 2.
\item No user response model. Burden accumulation and 4 archetypes are Phase 2.
\item Transition matrix is fixed and known. Learned transitions from real data are Phase 2.
\item Transition probabilities are placeholder values informed by Reactance theory. Calibration against real data is Phase 2.
\end{itemize}

\end{document}
