\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,booktabs,graphicx}
\usepackage[margin=1in]{geometry}

\title{Experiment 1a: 6-Action Extension}
\author{William Dennis}
\date{June 2026}

\begin{document}
\maketitle

\section{Motivation}

The MVP simulator uses a binary action space (\texttt{nudge}, \texttt{idle}).
Issue \#126 extends this to 6 clinically-motivated actions with per-action
metadata (\texttt{reward\_penalty}, \texttt{burden\_penalty}) while maintaining
full backward compatibility with \texttt{list[str]} configs.

\section{Config Schema}

Actions now accept two formats:

\begin{itemize}
\item \texttt{list[str]}: MVP mode, e.g.\ \texttt{[nudge, idle]}
\item \texttt{list[dict]}: Extended mode with metadata:
\begin{verbatim}
- name: motivational_prompt
  reward_penalty: 0.1
  burden_penalty: 0.2
\end{verbatim}
\end{itemize}

The \texttt{\_normalize\_actions()} validator extracts names, rejects mixed
types and duplicates. Agents receive the flat \texttt{action\_names} list.

\section{6-Action Config}

\begin{center}
\begin{tabular}{lcc}
\toprule
Action & Reward Penalty & Burden Penalty \\
\midrule
no\_message & 0.00 & 0.00 \\
motivational\_prompt & 0.10 & 0.20 \\
walking\_suggestion & 0.20 & 0.30 \\
goal\_reminder & 0.05 & 0.15 \\
recovery\_suggestion & 0.15 & 0.25 \\
progress\_feedback & 0.08 & 0.10 \\
\bottomrule
\end{tabular}
\end{center}

\section{Results (50 seeds, 90 days)}

\begin{center}
\begin{tabular}{lrrr}
\toprule
Agent & Total Reward & Per Step \\
\midrule
Standard TS & 260.8 $\pm$ 23.8 & 0.5796 \\
Contextual TS & 248.8 $\pm$ 21.2 & 0.5529 \\
Standard EG & 260.1 $\pm$ 33.0 & 0.5781 \\
Contextual EG & 249.4 $\pm$ 30.8 & 0.5543 \\
Standard UCB & 270.4 $\pm$ 24.5 & 0.6009 \\
Contextual UCB & 254.4 $\pm$ 19.6 & 0.5653 \\
Standard DEC & 268.6 $\pm$ 32.5 & 0.5969 \\
Contextual DEC & 255.7 $\pm$ 25.0 & 0.5681 \\
Random & 211.4 $\pm$ 17.1 & 0.4699 \\
\bottomrule
\end{tabular}
\end{center}

Standard UCB achieves the highest reward (270.4). Contextual agents do not
outperform standard variants on this config, likely because the 6-action
space is too large for activity-only conditioning to learn effectively
within 450 steps.

\section{Open Questions}

\begin{itemize}
\item The penalty values are hand-tuned placeholders. A grid search over
penalty combinations could identify regions where burden penalty meaningfully
shapes agent behaviour.
\item With 6 actions and only 2 states, the transition matrix has 12 entries.
Real clinical data would inform more realistic transition probabilities.
\end{itemize}

\end{document}
