\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,booktabs,graphicx}
\usepackage[margin=1in]{geometry}

\title{Initial Experiments: Extending the MVP Simulator}
\author{William Dennis}
\date{June 2026}

\begin{document}
\maketitle

Extends the MVP simulator (\texttt{docs/mvp/mvp\_specification.tex}) by enriching
the state space, action space, and reward function, moving toward the full
MDP defined in \texttt{docs/design/initial\_design.tex}.

\section{State Space Extension}
\label{sec:state}

References \texttt{docs/design/initial\_design.tex \S3} and Issue \#124.
\emph{Implementation: PR 1b (feat/1b-states) --- merged.}


\subsection{Enriched StateView}

The MVP dataclass currently holds four fields:
\[
\texttt{StateView(activity, day, step\_of\_day, steps\_per\_day)}
\]

PR~1b adds four optional float fields (all defaulting to \texttt{None} for
MVP backward compatibility):
\[
\texttt{steps},\quad \texttt{weight},\quad \texttt{time\_of\_day},\quad \texttt{day\_of\_week}
\]

The transition ABC signature changes from
$\texttt{transition(state: str, action: str) -> str}$ to
$\texttt{transition(state: StateView, action: str) -> StateView}$.
The reward ABC signature changes from $\texttt{reward(state: str, ...)}$ to
$\texttt{reward(state: StateView, ...)}$.

When the config includes a \texttt{state\_dynamics} block,
\texttt{RuleBasedTransition} enters extended mode and evolves steps and weight
according to the following equations:

\begin{align}
\text{tod} &= \text{step\_of\_day} \times \frac{24}{\text{steps\_per\_day}} \\
\text{dow} &= \text{day} \bmod 7 \\
\text{steps\_mean} &= \text{response\_multiplier}[a] \times
\text{tod\_modulation}[\text{tod}] \times
\text{dow\_modulation}[\text{dow}] \\
\text{steps}_{t+1} &= \max\bigl(0,\; \text{steps}_t + \text{steps\_mean} +
\mathcal{N}(0, \sigma_{\text{steps}})\bigr) \\
\Delta\text{steps} &= \text{steps}_{t+1} - \text{steps}_t \\
\text{weight}_{t+1} &= \text{weight}_t + \text{meal\_effect}[\text{tod}] +
\text{weekend\_boost} \cdot \mathbb{1}_{\text{dow} \in \{0, 6\}} +
\text{steps\_coefficient} \cdot \Delta\text{steps} +
\mathcal{N}(0, \sigma_{\text{weight}})
\end{align}

The config schema defines two Pydantic models:
\begin{itemize}
\item \texttt{StepsDynamicsConfig}: \texttt{response\_multiplier} (per-action step
response), \texttt{tod\_modulation} (hour-of-day gain, default 1.0 for missing
hours), \texttt{dow\_modulation} (day-of-week gain, default 1.0), and
\texttt{noise\_std} (non-negative, default 50).
\item \texttt{WeightDynamicsConfig}: \texttt{meal\_effect} (per-hour weight
change), \texttt{weekend\_boost} (additional weight gain on Sat/Sun, default
0.05), \texttt{steps\_coefficient} (weight loss per step, default $-10^{-4}$),
and \texttt{noise\_std} (non-negative, default 0.05).
\end{itemize}

The experiment config is at
\texttt{docs/initial\_experiments/configs/1b\_states\_only.yaml}.
It uses the 2 MVP actions with the above dynamics.

\subsection{Environment Evolution (PR 1b)}

\texttt{Environment.reset()} initialises \texttt{steps} and \texttt{weight}
from the config's \texttt{initial\_steps} (default 0.0) and
\texttt{initial\_weight} (default 70.0). \texttt{time\_of\_day} and
\texttt{day\_of\_week} are computed deterministically from the step counter.

After each \texttt{step()} call, the environment overwrites the deterministic
calendar fields (\texttt{day}, \texttt{step\_of\_day}, \texttt{time\_of\_day},
\texttt{day\_of\_week}) via \texttt{dataclasses.replace}. The transition model
does not set these fields --- they are the environment's responsibility.

\subsection{Open Questions}

\begin{itemize}
\item The \texttt{tod\_modulation} keys are sparse ($\{0,6,12,18,23\}$).
Missing hours default to~1.0, meaning most steps receive no time-of-day
modulation. This may be refined with higher-resolution data.
\item Steps and weight are unnormalised. Normalisation may be needed when these
fields are used as agent context features.
\end{itemize}


\section{Action Space Extension}
\label{sec:action}

References \texttt{docs/design/initial\_design.tex \S3} and Issue \#126.


\subsection{ActionConfig}

% To be filled: reward_penalty, burden_penalty, 6-action set


\subsection{Transition Model Updates}

% To be filled: 6x6x2 transition matrix, backward compatibility


\subsection{Agent Compatibility}

% To be filled: verification of variable-length action handling


\subsection{Open Questions}

% To be filled


\section{Reward Function Extension}
\label{sec:reward}

References \texttt{docs/design/initial\_design.tex \S3} and Issue \#132.


\subsection{Multi-Timescale Reward}

% To be filled: immediate + 3-week delayed clinical outcome


\subsection{Reward Composition}

% To be filled: configurable per_step_weight, clinical_weight


\subsection{Synthetic Clinical Data}

% To be filled: weekly_step_count, engagement_ratio generator


\subsection{Open Questions}

% To be filled


\section{Combined Results}

Results from the pairwise and full-integration experiments will appear here
once all three extension PRs (1a, 1b, 1c) have been merged. Currently available:

\begin{itemize}
\item \texttt{docs/initial\_experiments/configs/1b\_states\_only.yaml} (PR~1b,
2~actions, 4~continuous states, simple reward) --- see \S\ref{sec:state}.
\end{itemize}

\paragraph{PR~1b Results (50 seeds, all MVP agents).}

\begin{center}
\begin{tabular}{lrrrr}
\toprule
Agent & Total Reward & Per Step & Weight (kg) & Steps \\
\midrule
Standard TS & 161.8 $\pm$ 15.9 & 0.3596 & 68.57 $\pm$ 1.22 & 57977 $\pm$ 5667 \\
Contextual TS & 178.8 $\pm$ 15.8 & 0.3974 & 69.50 $\pm$ 1.06 & 48702 $\pm$ 5210 \\
Standard EG & 155.3 $\pm$ 29.0 & 0.3452 & 68.80 $\pm$ 1.82 & 55728 $\pm$ 14096 \\
Contextual EG & 172.7 $\pm$ 26.9 & 0.3837 & 69.63 $\pm$ 1.62 & 47355 $\pm$ 10842 \\
Standard UCB & 163.6 $\pm$ 15.6 & 0.3635 & 68.34 $\pm$ 1.19 & 60253 $\pm$ 5517 \\
Contextual UCB & 179.2 $\pm$ 17.4 & 0.3982 & 69.51 $\pm$ 1.16 & 48626 $\pm$ 6131 \\
Standard DEC & 157.2 $\pm$ 25.6 & 0.3494 & 68.70 $\pm$ 1.70 & 56673 $\pm$ 13265 \\
Contextual DEC & 177.0 $\pm$ 28.8 & 0.3934 & 69.64 $\pm$ 1.45 & 47318 $\pm$ 9436 \\
\bottomrule
\end{tabular}
\end{center}\vspace{2mm}

Over 90-day episodes (450 steps) across 50 seeds, contextual agents
consistently outperform their non-contextual counterparts by $\sim$10--15
reward points. Contextual UCB achieves the highest mean reward
($179.2 \pm 17.4$) while contextual TS and DEC follow closely. Weight
drops from 70.0~kg across all agents ($68.3$--$69.6$~kg), driven by the
cumulative meal effects and the negative \texttt{steps\_coefficient}.
Total steps range from 47k to 60k; contextual agents take fewer steps
(47k--49k) yet achieve higher rewards, suggesting they learn to prefer
\texttt{nudge} at higher-modulation hours. The per-step reward gap
between contextual ($\sim$0.39) and non-contextual ($\sim$0.35) agents
confirms that activity-based conditioning improves policy quality even
with simple transition dynamics.

\end{document}
