\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{url}

\geometry{margin=2.5cm}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\title{Design Document: RL-Driven Physical Activity Interventions\\
\large A Configurable Simulation Framework}
\author{William Dennis\\
Bristol x NUS Research Project\\
\texttt{ky23812@bristol.ac.uk}}
\date{June 2026}

\begin{document}
\maketitle
\tableofcontents

\newpage

\section{Problem Statement}
\label{sec:problem}

\subsection{Background}

Physical inactivity is a leading risk factor for chronic disease worldwide.
Just-in-time adaptive interventions (JITAIs) delivered through wearable devices
aim to provide the right support at the right moment, leveraging real-time
sensor data to trigger personalised prompts. However, designing effective
intervention policies---deciding \emph{when}, \emph{what}, and \emph{how
much} to intervene---is a sequential decision-making problem that traditional
rule-based approaches handle poorly.

Reinforcement learning provides a natural framework for this problem: an agent
learns a policy that maps user state (step count, sleep, heart rate, time of
day) to an intervention action (motivational prompt, goal reminder, rest
suggestion), optimising long-term behaviour change rather than short-term
response. The key challenge is that training RL agents requires either
expensive real-world trials or a realistic simulation environment.

\subsection{Problem}

There is no standardised, configurable simulation platform for evaluating RL
agents on PA intervention tasks. Existing work uses bespoke simulators tied to
specific datasets or papers. This makes cross-paper comparison difficult and
limits reproducibility. Five specific gaps in the current landscape motivate
this work:

\begin{enumerate}[noitemsep]
    \item No LLM simulation validated for wearable health data
    \item No delayed-feedback bandit for 3-week clinical measures
    \item No non-stationary adaptation for PA interventions
    \item No standardised MDP/benchmark for PA JITAI RL \cite{karine2024stepcountjitai}
    \item No offline RL for PA interventions with real wearable data
\end{enumerate}

These gaps are \emph{experiments run on the framework}, not part of the
framework itself. The framework provides the infrastructure to investigate
each one.

\subsection{Goals}

This project delivers:

\begin{enumerate}[noitemsep]
    \item A \textbf{config-first simulation framework} where the dataset
    schema, MDP, user behaviour, and agents are specified entirely via YAML
    config files.
    \item A \textbf{configurable MDP environment} supporting multi-timescale
    rewards (daily steps + delayed clinical measures).
    \item A \textbf{rule-based user simulator} grounded in behavioural theory,
    with configurable archetypes and fatigue mechanisms.
    \item A \textbf{library of RL agents} sharing a common interface, from
    simple baselines (Thompson Sampling) to deep RL (DQN, PPO).
    \item An \textbf{experiment runner} for reproducible multi-trial,
    multi-agent comparisons.
\end{enumerate}

\subsection{Scope}

Phase 1 (8 weeks) delivers a working CLI: \texttt{uv run rl-health-interventions
--config experiment.yml} that trains agents and produces a results table.
Phase 2 (stretch) adds LLM-based user simulation and validation against real
data.

\section{Design Requirements}
\label{sec:requirements}

\subsection{Stakeholders}

\begin{itemize}[noitemsep]
    \item \textbf{William Dennis} -- implementation, single developer
    \item \textbf{Mengyan Zhang} (Bristol) -- domain expertise, data access,
    supervision
    \item \textbf{Swapnil Mishra} (NUS / MLGH) -- supervision, downstream
    research team
\end{itemize}

\subsection{Constraints}

\begin{itemize}[noitemsep]
    \item 8-week internship timeline
    \item Single developer, no prior codebase
    \item No existing real data access (4-8 week institutional applications
    required)
    \item Deliverable must be usable by Swapnil's team after handover
\end{itemize}

\subsection{Requirements}

\begin{description}[noitemsep]
    \item[R1 -- Config-first] \hfill \\
    All components (dataset schema, MDP, user behaviour, agents, experiments)
    are specified via YAML config files. New dataset or intervention design
    requires a new config, not new code.

    \item[R2 -- Reproducible] \hfill \\
    Fixed seeds, config snapshot captured per run, deterministic evaluation
    mode.

    \item[R3 -- Pluggable] \hfill \\
    Each component family (transitions, rewards, agents, response models) uses
    an ABC + registry pattern. Adding a new implementation requires one file
    and one registry entry.

    \item[R4 -- Minimal dependencies] \hfill \\
    Python 3.11, \texttt{uv}, Polars, Pydantic, NumPy. No Stable Baselines or
    Gymnasium unless proven necessary.

    \item[R5 -- Production-quality error handling] \hfill \\
    Three-layer validation (schema, compatibility, dummy step). Per-episode
    exception isolation. Structured JSON trace output.

    \item[R6 -- Single-entry experiment] \hfill \\
    \texttt{uv run rl-health-interventions --config experiment.yml} builds,
    runs, and reports. No multi-step workflow.
\end{description}

\subsection{Non-Requirements}

The following are explicitly out of scope for Phase 1:

\begin{itemize}[noitemsep]
    \item Web UI or dashboard
    \item Real data integration (All of Us, UK Biobank)
    \item Deep RL agents (DQN, PPO, CQL, IQL)
    \item LLM-based user simulation
    \item Gymnasium or SB3 compatibility layer
    \item Real-time or online learning
\end{itemize}

\section{Architecture Overview}
\label{sec:architecture}

\subsection{Design Principles}

\begin{itemize}[noitemsep]
    \item \textbf{Config-first:} Every component reads from YAML. New dataset
    or intervention design = new config, not new code.
    \item \textbf{Pluggable components:} Each family (transitions, rewards,
    agents, response models) uses an ABC + registry pattern. Adding a new
    component requires one file, one registry entry, and a config string
    reference.
    \item \textbf{Minimal dependencies:} Python 3.11, \texttt{uv}, Polars,
    Pydantic, NumPy. No Stable Baselines or Gymnasium unless proven necessary.
    \item \textbf{Reproducibility:} Config snapshots, fixed seeds, deterministic
    evaluation.
\end{itemize}

\subsection{Component Overview}

The framework is organised into five subphases, each delivering a pluggable
component:

\begin{enumerate}[noitemsep]
    \item \textbf{Data Layer (1A):} Ingests CSV/Parquet files via Polars lazy
    scans, applies config-driven feature engineering, and materialises a
    numpy-backed \texttt{Dataset}. Provides a \texttt{StateView} bridge for
    the simulation loop.

    \item \textbf{MDP Environment (1B):} Exposes a clean
    \texttt{step(state, action) $\rightarrow$ (next\_state, reward, done)} API.
    Pluggable \texttt{TransitionModel} and \texttt{RewardHandler} components
    are selected from a registry by config string.

    \item \textbf{User Simulation (1C):} Rule-based behavioural archetypes
    (goal-driven, social responder, resistant, stable maintainer) with
    backlash and fatigue mechanisms. A \texttt{ResponseModel} maps
    $(\text{state}, \text{action}, \text{profile})$ to a next state.

    \item \textbf{Agent Library (1D):} Common \texttt{Agent} ABC with
    \texttt{select\_action(state) $\rightarrow$ int} and
    \texttt{update(...)}. Thompson Sampling is the initial implementation;
    deep RL agents (DQN, PPO, CQL, IQL) are stretch.

    \item \textbf{Experiment Runner (1E):} Wires all components via a single
    \texttt{ExperimentFactory.build(config)}. Runs multi-trial, multi-agent
    experiments with config snapshot reproducibility.
\end{enumerate}

\subsection{Data Pipeline}

\begin{verbatim}
Raw files (CSV, Parquet, ...)      <- researcher provides these
    |
Polars LazyFrame scan               <- config-driven schema mapping
    |
FeaturePipeline.from_config(config) <- config-driven transforms (agg, window, diff, etc.)
    |  (Polars lazy graph, materialised once)
numpy arrays                         <- materialised for sim loop
    |
Dataset (user x time index)          <- random access for StateView slices
\end{verbatim}

Polars owns the feature engineering stage. Its lazy API maps one-to-one with
the config-driven specification---grouping, rolling windows, expressions. The
framework builds a \texttt{pl.LazyFrame} from the config and materialises it
once. The result is converted to numpy arrays for the simulation loop, which
needs fast random access by user and time rather than query capability.

After materialisation, the simulation depends only on numpy---no Polars
dependency in the hot path.

\subsection{Module Layout}

Each pluggable concept gets a flat package with a \texttt{\_base.py} ABC and a
\texttt{REGISTRY} dict in \texttt{\_\_init\_\_.py}:

\begin{verbatim}
transitions/               <- pluggable transition models
    _base.py               <- TransitionModel ABC
    rule_based.py          <- RuleBasedTransition
    __init__.py            <- imports + REGISTRY dict
rewards/                   <- pluggable reward functions
    _base.py               <- RewardHandler ABC
    compound.py            <- CompoundReward
    __init__.py            <- imports + REGISTRY dict
agents/                    <- pluggable agents
    _base.py               <- Agent ABC
    thompson_sampling.py   <- ThompsonSamplingAgent
    __init__.py            <- imports + REGISTRY dict
simulation/                <- pluggable user response models
    _base.py               <- ResponseModel ABC
    rule_based.py          <- RuleBasedResponse
    __init__.py            <- imports + REGISTRY dict
data/                      <- data layer
    _base.py               <- IngestPipeline ABC
    feature_pipeline.py    <- Config-driven feature engineering on Polars
    polars_reader.py       <- Reads CSV/Parquet via Polars lazy API
    dataset.py             <- Dataset (numpy arrays, for sim loop)
    synthetic.py           <- SyntheticDataGenerator
    __init__.py            <- imports + REGISTRY dict
\end{verbatim}

Rules:
\begin{itemize}[noitemsep]
    \item \texttt{\_base.py} contains only the ABC. No implementations.
    \item Each implementation is its own file, named after the class in
    snake\_case.
    \item \texttt{\_\_init\_\_.py} explicitly imports every implementation and
    populates a \texttt{REGISTRY} dict. No auto-discovery.
\end{itemize}

\subsection{Configurability}

All MDP components are exposed in the config schema:

\begin{table}[h]
\centering
\caption{Configurable parameters}
\begin{tabular}{lll}
\toprule
Category & Parameter & Default \\
\midrule
State & Variable selection & All \\
Actions & Available subsets & $\{a_0, \dots, a_5\}$ \\
Reward & $(\alpha, \beta, \lambda, \eta)$ & $(1.0, 0.1, 0.5, 1.0)$ \\
Transition & Archetype & goal-driven \\
Fatigue & $B_{\max}$ & 3 \\
Discount & $\gamma$ & 0.95 \\
Epoch & Frequency & Daily \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Validation}

The factory applies three layers of validation before the main experiment loop:

\begin{enumerate}[noitemsep]
    \item \textbf{Schema validation (Pydantic):} Malformed configs are rejected
    at load time with a precise error message.
    \item \textbf{Component compatibility:} Registry keys and cross-references
    are checked before instantiation.
    \item \textbf{Dummy step:} A single \texttt{step()} call catches wiring
    errors and shape mismatches before the main loop.
\end{enumerate}

\subsection{Test Organisation}

\begin{verbatim}
tests/
    unit/
        transitions/       <- each transition model in isolation
        rewards/           <- each reward function in isolation
        agents/            <- each agent in isolation
        simulation/        <- each response model in isolation
        factory/           <- factory validation and wiring
    integration/
        test_dummy_step.py <- factory builds and step succeeds
        test_config.yml    <- minimal config is loadable
    end_to_end/
        test_experiment.py <- full pipeline: config -> build -> train -> results
\end{verbatim}

Rules:
\begin{itemize}[noitemsep]
    \item Every ABC method has at least one test with a real implementation.
    \item Every subphase gate requires all tests in that subphase's modules.
    \item Test structure maps to modules, not subphase numbers.
    \item Registry discovery test ensures every registered component can be
    instantiated from a minimal config.
\end{itemize}

\section{Decision Log}
\label{sec:decisions}

\begin{table}[h]
\centering
\caption{Design decisions}
\begin{tabular}{cll}
\toprule
\# & Decision & Status \\
\midrule
1 & Config format: YAML & Confirmed \\
2 & Interface: CLI + config files; web UI = stretch & Confirmed \\
3 & Python 3.11 + \texttt{uv} for package management & Confirmed \\
4 & Synthetic data for Phase 1 (real data requires 4-8 week applications) & Confirmed \\
5 & Thompson Sampling as initial agent & Confirmed \\
6 & Daily decision epochs & Confirmed \\
7 & Multi-timescale reward: immediate (daily steps) + delayed (3-week body measures) & Confirmed \\
8 & Four behavioural archetypes: goal-driven, social responder, resistant, stable maintainer & Confirmed \\
9 & Pluggable component architecture (ABC + registry pattern) & Confirmed \\
10 & Minimal dependencies: no Gymnasium, no SB3 & Confirmed \\
\bottomrule
\end{tabular}
\end{table}

\section{Open Decisions}
\label{sec:open}

The following decisions require supervisor input before they affect
implementation direction.

\subsection{Distribution model}

Should the framework be distributed as an installable package or remain
repo-based?

\begin{description}[noitemsep]
    \item[Option A -- Installable package] \hfill \\
    \texttt{uv build} produces a wheel. Researchers run
    \texttt{uv add rl-health-interventions} in their own project.
    Clearer dependency boundary, but requires CI/CD for publishing.

    \item[Option B -- Repo-based] \hfill \\
    Clone and \texttt{uv sync}. Simpler for a single-team prototype.
    Swapnil's team would fork or clone directly.
\end{description}

\subsection{MDP baseline approval}

Does the proposed MDP formalisation (\S6) look like a reasonable starting
point? The state space, action set, and reward function are intentionally
conservative---they can be extended via config without breaking changes.

\subsection{Phase 2 dataset targets}

Which datasets should we target for real-data validation?

A recommended order: HeartSteps first (fastest access, most relevant), then
All of Us (largest population), then UK Biobank (raw accelerometer).

\subsection{Simulator validation criteria}

How do we determine the simulator is ``good enough'' before running
experiments? Candidate metrics include response-rate RMSE against real
MRT data, state-distribution KL divergence, and engagement-curve alignment.
We need agreement on acceptable thresholds.

\section{MDP Formalisation}
\label{sec:mdp}

This section specifies the Markov Decision Process (MDP) for the first
iteration of the simulation framework. The MDP is defined as the tuple
$(\mathcal{S}, \mathcal{A}, P, R, \gamma)$.

\subsection{State Space $\mathcal{S}$}

The state represents the current context of a user at decision epoch $t$.
States are constructed from wearable-device signals, user profile
information, and intervention history:

\[
s_t = \{ \mathbf{x}_t^{\text{wearable}}, \mathbf{x}_t^{\text{context}},
\mathbf{x}_t^{\text{history}}, \mathbf{x}^{\text{profile}} \}
\]

\begin{table}[h]
\centering
\caption{State variables}
\begin{tabular}{lll}
\toprule
Variable & Type & Description \\
\midrule
$\text{steps}_t$ & $\mathbb{R}_{\geq 0}$ & Step count, past 24h \\
$\text{hr}_t$ & $\mathbb{R}_{\geq 0}$ & Heart rate (or resting) \\
$\text{sleep\_hours}_t$ & $\mathbb{R}_{\geq 0}$ & Previous night sleep \\
$\text{sedentary\_min}_t$ & $\mathbb{R}_{\geq 0}$ & Sedentary minutes today \\
$\text{time\_of\_day}_t$ & $\{0,\dots,23\}$ & Current hour \\
$\text{day\_of\_week}_t$ & $\{0,\dots,6\}$ & Day of week \\
$\text{goal\_progress}_t$ & $[0,1]$ & Fraction of daily goal met \\
$\text{burden}_t$ & $\mathbb{Z}_{\geq 0}$ & Interventions sent today \\
$a_{t-1}$ & $\mathcal{A}$ & Previous action \\
$\text{response}_{t-1}$ & $\{0,1\}$ & Previous response \\
$\text{body\_measure}_k$ & $\mathbb{R}$ & Clinical measure (q3wks) \\
$\text{age}$ & $\mathbb{Z}_{>0}$ & Static profile \\
$\text{gender}$ & $\{0,1\}$ & Static profile \\
$\text{baseline\_activity}$ & $\{0,1,2\}$ & Low / medium / high \\
\bottomrule
\end{tabular}
\end{table}

Variables are normalised to $[0,1]$ before being passed to the agent. The
body measure is only observed every 3 weeks (sparse reward signal).

\subsection{Action Space $\mathcal{A}$}

\[
\mathcal{A} = \{ a_0, a_1, a_2, a_3, a_4, a_5 \}
\]

\begin{table}[h]
\centering
\caption{Action space}
\begin{tabular}{cll}
\toprule
Action & Label & Description \\
\midrule
$a_0$ & No message & Do not intervene \\
$a_1$ & Motivational prompt & Generic encouragement \\
$a_2$ & Walking suggestion & Recommend a short walk \\
$a_3$ & Goal reminder & Remind of step goal \\
$a_4$ & Recovery suggestion & Stretch or rest \\
$a_5$ & Progress feedback & Report goal progress \\
\bottomrule
\end{tabular}
\end{table}

Exactly one action per decision epoch. The ``no message'' action is critical
for managing notification burden.

\subsection{Transition Function $P$}

The transition function models the probability over next states:

\[
P(s_{t+1} \mid s_t, a_t): \mathcal{S} \times \mathcal{A} \times \mathcal{S}
\rightarrow [0,1]
\]

In Phase 1 (rule-based), transitions follow a behavioural response model:

\[
\Delta \text{steps}_{t+1} = f(s_t, a_t, \theta_{\text{user}})
\]

where $\theta_{\text{user}}$ parameterises one of four behavioural archetypes:

\begin{itemize}[noitemsep]
    \item \textbf{Goal-driven:} Responds to reminders and feedback.
    Proportional to goal progress.
    \item \textbf{Social responder:} Responds to motivational prompts.
    Models social desirability bias.
    \item \textbf{Resistant:} Low response. Fatigue accumulates quickly.
    \item \textbf{Stable maintainer:} Already active. Low marginal gain.
\end{itemize}

Burden evolves as:

\[
\text{burden}_{t+1} =
\begin{cases}
\text{burden}_t + 1 & \text{if } a_t \neq a_0 \\
0 & \text{if } a_t = a_0
\end{cases}
\]

When $\text{burden} > B_{\max}$, response probability decays linearly.

\subsection{Reward Function $R$}
\label{sec:reward}

\subsubsection{Immediate Reward}

\[
R_t^{\text{immediate}} = \alpha \cdot \Delta\text{steps}_t
- \beta \cdot \mathbf{1}_{\{a_t \neq a_0\}}
+ \lambda \cdot \text{goal\_progress}_t
\]

\subsubsection{Delayed Reward (Body Measures)}

Clinical body measures arrive every 3 weeks ($t = 3k$):

\[
R_{3k}^{\text{delayed}} = \eta \cdot \text{BM\_improvement}_{3k}
\]

Total reward:

\[
R_t =
\begin{cases}
R_t^{\text{immediate}} + R_t^{\text{delayed}} & t \equiv 0 \pmod{3} \\
R_t^{\text{immediate}} & \text{otherwise}
\end{cases}
\]

The compound reward (immediate + delayed) is a central research question.
Weights $(\alpha, \beta, \lambda, \eta)$ are configurable experiment parameters.

\subsection{Discount Factor $\gamma$}

\[
\gamma \in [0.9, 0.95]
\]

A high discount factor encourages long-term behaviour change over short-term
step increases. Appropriate for health habit formation.

\subsection{Decision Epoch}

\textbf{Daily} decisions. This aligns with typical intervention cadence,
the 3-week body measure delay (3 epochs), and practical deployment
constraints.

\section{Simulator Training Method}
\label{sec:sim}

The user simulator (1C) transforms a current state and chosen action into
a next state via a parameterised response model:

\[
s_{t+1} = \text{ResponseModel}(s_t, a_t; \theta)
\]

This section describes how $\theta$ is estimated and validated.

\subsection{Parameter Estimation from MRT Data}

Micro-randomized trials (HeartSteps V1/V2) provide sequences of
$(s_t, a_t, s_{t+1})$ tuples. For each user, we observe how their step
count changes after each intervention type. The parameter vector
$\theta$ governs:

\begin{itemize}[noitemsep]
    \item \textbf{Baseline step increase:} Expected $\Delta$steps when
    no intervention is sent
    \item \textbf{Intervention effect size:} Marginal $\Delta$steps per
    action type, modelled as $\Delta_{\text{effect}}(a_t) \sim
    \mathcal{N}(\mu_a, \sigma_a^2)$
    \item \textbf{Fatigue sensitivity:} Decay rate $\delta$ such that
    response probability $p(\text{respond}) = \max(0, 1 - \delta \cdot
    \text{burden})$
    \item \textbf{Burden threshold:} $B_{\max}$, the number of consecutive
    interventions before response probability reaches zero
\end{itemize}

Estimation procedure:

\begin{enumerate}[noitemsep]
    \item Fit baseline step distribution from non-intervention control
    periods
    \item Compute empirical effect sizes per action type from intervention
    windows
    \item Cluster users by response pattern to assign archetype labels
    \item Estimate archetype-specific parameters via maximum likelihood
    on held-out users
\end{enumerate}

In Phase 1, synthetic parameter values are drawn from published statistics
(mean step counts by age and gender, published HeartSteps engagement
curves) \cite{klasnja2019heartsteps,klasnja2022heartsteps}. Real MRT data calibration is Phase 2.

\subsection{Calibration Loop}

After parameter estimation, the simulator is calibrated by comparing
population-level statistics:

\begin{enumerate}[noitemsep]
    \item Generate $N$ synthetic user trajectories from the simulator
    \item Compute target statistics from real data (or published
    benchmarks): mean daily steps, response rate by week, burden
    accumulation rate
    \item Adjust $\theta$ to minimise discrepancy between simulated and
    target statistics
    \item Repeat until convergence or budget exhausted
\end{enumerate}

\subsection{Fidelity Metrics}

\begin{description}[noitemsep]
    \item[Response rate RMSE] \hfill \\
    Root mean squared error between simulated and real response rates,
    bucketed by week and action type. Target: $<0.05$.

    \item[Step count distribution KL divergence] \hfill \\
    KL divergence between simulated and real daily step distributions.
    Target: $<0.1$ nats.

    \item[Engagement decay alignment] \hfill \\
    Slope of engagement-over-time curve. Target: within 20\% of real
    decay rate.

    \item[Archetype separation] \hfill \\
    Do distinct archetypes produce distinct reward trajectories?
    Measured by inter-archetype mean reward variance.
\end{description}

\subsection{Limitations}

The rule-based simulators in Phase 1 capture first-order effects only:
\begin{itemize}[noitemsep]
    \item No habituation or context-dependent preference shifts
    \item No social or environmental confounders (weather, holidays)
    \item No long-term habit formation (positive reinforcement compound
    effects)
\end{itemize}

These limitations motivate Phase 2 LLM-based simulation.

\section{Data Sources}
\label{sec:data}

This project targets two tiers of data: synthetic for Phase 1
development and real-world for Phase 2 validation.

\subsection{Phase 1: Synthetic Data}

All Phase 1 experiments use synthetic data generators parameterised from
published statistics. This avoids the 4-8 week lead time for real dataset
access. The \texttt{SyntheticDataGenerator} produces realistic wearable
traces by sampling from population distributions reported in the literature
(e.g., NHANES step counts by age/gender \cite{nhanes2025steps}, All of Us sleep distributions \cite{patten2026allofus}).

\subsection{Phase 2: Real Data Targets}

\begin{enumerate}[noitemsep]
    \item \textbf{All of Us Fitbit Dataset} \cite{patten2026allofus}: 59,000+
    participants, Fitbit data, 14-year span. Access via cloud workbench
    only.

    \item \textbf{UK Biobank Accelerometer Dataset} \cite{williamson2024ukbiobank}:
    700,000+ person-days, wrist-worn tri-axial accelerometer.
    Requires institutional application (4-8 weeks).

    \item \textbf{HeartSteps V1/V2}: Micro-randomized trials with
    intervention response data. Directly calibrates the user simulation
    (1C). Access requires data sharing agreement (weeks to months).
\end{enumerate}

\subsection{Dataset Survey}

See \texttt{docs/sources/data\_sources.md} for the full feasibility study
(All of Us, UK Biobank) and \texttt{docs/sources/additional\_data\_sources.md}
for a survey of 12 additional datasets including HeartSteps, WISDM,
ExtraSensory, and open-access coaching intervention datasets. Key findings:

\begin{itemize}[noitemsep]
    \item Between All of Us and UK Biobank, most wearable MDP variables
    (steps, HR, sleep, sedentary, time-of-day) are covered.
    \item Gaps are simulated variables: goal progress, burden, user
    archetype.
    \item Body measures need EHR linkage (All of Us has 46\% linked).
    \item HeartSteps is the only available source of action-to-response
    mappings for direct simulator calibration.
\end{itemize}

\section{Timeline \& Future Work}
\label{sec:timeline}

\subsection{Phase 1 Milestones}

\begin{table}[h]
\centering
\caption{Phase 1 milestones}
\begin{tabular}{ll}
\toprule
Week & Milestone \\
\midrule
1--2 & Data layer (1A) + MDP environment (1B) \\
3--4 & User simulation (1C) + Agent library (1D) \\
5--6 & Experiment runner (1E) + end-to-end tests \\
7--8 & Documentation, examples, handover \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Phase 2: Future Work}

\begin{itemize}[noitemsep]
    \item LLM-based user simulation (prompt-driven persona modelling)
    \item Offline RL agents (CQL, IQL) for batch policy learning
    \item Non-stationary environment adaptation
    \item Cross-dataset transfer experiments
    \item Real-data validation pipeline (All of Us, UK Biobank)
    \item Integration with real-world JITAIs
\end{itemize}

\bibliographystyle{unsrt}
\bibliography{references}

\end{document}
