\documentclass[10pt,a4paper]{article}

\usepackage[margin=1.7cm]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{titlesec}
\usepackage{parskip}

\setlength{\parskip}{3pt plus 1pt}
\titlespacing\section{0pt}{8pt plus 2pt}{4pt plus 2pt}
\titlespacing\subsection{0pt}{6pt plus 2pt}{2pt plus 2pt}

\title{\vspace{-1cm}Sprint 1 Handover Notes\\{\large Config-Driven RL for Just-in-Time Adaptive Interventions}}
\author{William Dennis --- University of Bristol \& National University of Singapore}
\date{July 2026}

\begin{document}
\maketitle
\thispagestyle{empty}

\section{Problem Definition}

Physical inactivity is a leading global health risk. Just-in-Time Adaptive Interventions (JITAIs) deliver personalised behaviour-change prompts via wearable devices at moments of opportunity~\cite{klasnja2019heartsteps}. Designing effective JITAI policies requires rapid iteration, but real-world trials are slow, expensive, and cannot explore counterfactual policies.

We formulate the problem as an MDP \(\mathcal{M} = (\mathcal{S}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma)\):
\begin{itemize}
  \item \(\mathcal{S}\): 36-state factored space (step\_bin \(\in\) \{inactive, moderate, active\}, sleep \(\in\) \{good, poor\}, day\_of\_week \(\in\) \{weekday, weekend\}, burden \(\in\) \{low, medium, high\})
  \item \(\mathcal{A}\): 4 actions (idle, movement\_suggestion, goal\_reminder, journal)
  \item \(\mathcal{R} = \alpha \cdot f(\text{step\_bin}) + (1 - \alpha) \cdot g(\text{sleep}) - \lambda \cdot \mathbb{I}[\text{action} \neq \text{idle}]\)
  \item 5 decision points per day, 90-day default episode, 50 seeds per agent
\end{itemize}

\section{Existing Work}
\vspace{-2pt}
\paragraph{HeartSteps V1/V2} (Klasnja et al., 2019~\cite{klasnja2019heartsteps}; Klasnja et al., 2022~\cite{klasnja2022heartsteps}) --- First RL-enabled MRT for physical activity. V1 ATE +35 steps; V2 introduced Thompson Sampling with dosage tracking and proxy value function for delayed effects.

\paragraph{PEARL} (Lee et al., 2025~\cite{lee2025pearl}) --- Largest RL-for-PA RCT (n=13,463, 4-arm). COM-B informed 155-nudge bank. RL +296 steps vs control (p=0.0002) at 1 month, sustained at 2 months.

\paragraph{StepCountJITAI} (Karine \& Marlin, 2024~\cite{karine2024stepcountjitai}) --- Simulator for PA interventions, DQN $\approx$3,000 return vs TS $\approx$1,500.

\paragraph{Bandit theory} Russo \& Van Roy (2018)~\cite{russo2018learning} establish TS regret bound \(O(\sqrt{dT\log T})\). Trella et al. (2022)~\cite{trella2022designing} provide design principles for RL in digital health.

\paragraph{LLM-based simulation} Park et al. (2023)~\cite{park2023generative} demonstrate generative agents; Wang et al. (2024) use LLMs to simulate patient responses; Lu et al. (2025) benchmark LLM agents for multi-turn human behaviour.

\paragraph{Data-Driven Simulators \& Burden} Wang et al. (2021)~\cite{wang2021optimizing} developed behavioral simulators incorporating hard frequency constraints to mitigate over-notification and habituation.

\paragraph{User Heterogeneity} Tomkins et al. (2021)~\cite{tomkins2021intelligentpooling} introduced IntelligentPooling to balance personalization with group-level data sharing via mixed-effects Thompson Sampling. Ghosh et al. (2024)~\cite{ghosh2024rebandit} extended this with reBandit (Empirical Bayes + informative priors).

\paragraph{Count-Based Bandits} Liu et al. (2025)~\cite{liu2025thompson} extended TS to Poisson/NB/ZIP/ZINB models, reducing variance for skewed proximal health outcomes.

\paragraph{Power-Constrained Exploration} Yao et al. (2021)~\cite{yao2021power} established meta-algorithms for probability clipping (\(\pi_{min} \le P(A|S) \le \pi_{max}\)) to protect post-trial statistical power in MRTs.

\section{Our Approach}
\vspace{-2pt}
\textbf{Two pillars:} (1) config-driven framework defining entire MDPs in YAML --- no source code changes between experiments; (2) LLM-bootstrapped transition models that generate plausible patient behaviour from persona descriptions.

\textbf{Config-Driven Framework.} State spaces, action sets, transition models, reward functions, and agent selection are defined declaratively. Three transition types: rule\_based (hand-curated), random (Dirichlet baseline), and bootstrap (LLM-generated). Expression-based rewards via safe AST parser. Seven agents implemented with contextual variants, all pluggable by name in YAML.

\textbf{Synthetic LLM Patients.} Transition probability tables generated by prompting LLMs (DeepSeek V4 Flash) to sample next-state distributions given a persona description, current state, and action. Pipeline: prompt generation \(\rightarrow\) concurrent API calling (litellm, 200 workers) \(\rightarrow\) JSON parsing \(\rightarrow\) probability table aggregation. 22{,}320 prompts per persona (full factorial, N=10 per cell), 5 distinct behavioural personas, total cost \$0.27. 42 JSON transition tables, comprehensive analysis (parse rates, cell consistency, burden interactions, step histograms), 85 visualization figures.

\section{What We Have Done}
\vspace{-2pt}
\textbf{MVP Simulator.} A 2-state (sedentary/active), 2-action MDP with hand-curated transition rules. 285 hyperparameter combinations swept across epsilon-greedy, decaying epsilon-greedy, UCB, and Thompson Sampling variants. Full regression test suite with golden fixtures. Demonstrates the framework pipeline end-to-end.

\textbf{LLM Bootstrap.} Generated 42 JSON transition tables from 22{,}320 LLM calls (5 personas + 2 initial model comparisons). Each persona produces distinct, behaviourally-plausible transition dynamics validated through parse rates, output distributions, cell consistency, burden interaction analysis, and step count histograms. A 36-paper literature review across 8 dimensions supports the LLM-as-transition-model methodology.

\textbf{Experiments.} Both MDP types run through the same YAML schema, agent library, and evaluation runner. Benchmark matrix: 7 persona configurations \(\times\) 4 variant types \(\times\) 7 agents \(\times\) 50 seeds = 9,800 simulation runs. HeartSteps V2 reproduced faithfully (Bayesian linear regression, dosage tracking, proxy value function) as a validation agent --- 4 implementation bugs independently discovered during reproduction.

\section{Initial Results}

\textbf{Transition validation.} The LLM-generated personas produce meaningfully differentiated behaviour. Burden degrades movement suggestion effectiveness: as notification burden rises from low to high, transition probabilities to higher step bins decrease systematically. Persona contrast is pronounced: identical actions applied to different personas yield fundamentally different transition probability matrices (e.g., stable maintainers retain active states; resistant users drop to inactive regardless of intervention).

\textbf{Agent benchmarks.} HeartSteps V2 achieves 97--100\% of the optimal DP bound across all 7 persona/initial-model configurations, confirming the simulator is well-formed and can reproduce known results. Epsilon-greedy lags 10--40pp behind. Key findings:

\begin{table}[ht]
  \centering
  \small
  \begin{tabular}{lccc}
    \toprule
    Persona & Optimal DP & HeartSteps (\% opt) & Best Bandit (\% opt) \\
    \midrule
    Stable Maintainer & 441.8 & 440.8 (99.8\%) & 402.5 (91.1\%) \\
    Base                  & 379.7 & 379.1 (99.9\%) & 327.8 (86.3\%) \\
    Goal-driven           & 414.1 & 414.5 (100.1\%) & 323.0 (78.0\%) \\
    Social Responder      & 313.5 & 306.8 (97.9\%) & 140.1 (44.7\%) \\
    Initial DeepSeek      & 381.6 & 373.6 (97.9\%) & 297.5 (78.0\%) \\
    Initial GLM 5.2       & 374.6 & 374.0 (99.8\%) & 326.8 (87.2\%) \\
    Resistant             & 214.6 & 163.5 (76.2\%) & 151.7 (70.7\%) \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Observations.} (1) Persona effects are large --- Stable Maintainer (441) vs Resistant (215). (2) Epsilon-greedy is the best bandit but lags substantially behind HeartSteps. (3) HeartSteps selects idle $\approx$62\% of the time in the base persona --- personalisation is partly about knowing when \emph{not} to intervene. (4) Time-of-day features and step activity levels are the most predictive state variables.

\section{What is Next}

\begin{enumerate}
  \item \textbf{Sensor Foundation Models.} Wearable FM POC (2,754 lines, feature branch): NormWear integration (768-dim embeddings from PPG/accelerometer), FM-guided action space (15 interventions, 3 archetypes), Thompson Sampling with FM context keys. 6 publishable research gaps identified across 12 FM models.
  \item \textbf{Additional LLM Backends.} Extend bootstrapping to GLM 5.2, Qwen, Claude. Multi-turn persona dialogues for richer, temporally-aware transition models. Cost-benefit analysis across providers.
  \item \textbf{Better Prompting.} COM-B behaviour-change taxonomy integration. Structured JSON schema enforcement via constrained decoding. Chain-of-thought reasoning for transition probability generation.
  \item \textbf{Synthetic Clinical Results.} Calibrate LLM transitions to real-world effect sizes (PEARL: +200--300 steps). Integrate All of Us (n=59K, Fitbit) and UK Biobank (n=100K, accelerometer) for learned transition models.
  \item \textbf{Algorithmic advances.} Extend beyond bandits to DQN/policy-gradient for temporal credit assignment. Power-constrained exploration~\cite{yao2021power} and count-based likelihoods~\cite{liu2025thompson} for future MRT deployment.
\end{enumerate}

% --- References (not counted toward page limit) ---
\bibliographystyle{unsrtnat}
\bibliography{../design/references}

\end{document}

